---
title: "R Notebook"
output: html_notebook
---

```{r}
library(tidyverse)
library(janitor)
library(ggpubr)
library(broom)
library(AICcmodavg)

hospital_admissions <- read_csv("inpatient_and_daycase_by_nhs_board_of_treatment.csv")

hospital_admissions_speciality <- read_csv("inpatient_and_daycase_by_nhs_board_of_treatment_and_specialty.csv")
```

#cleaning
```{r}
#cleaning script 1
clean_hosp_admissions <- hospital_admissions %>% 
  clean_names() %>% 
  select(- quarter_qf, - hbqf, -location_qf, -admission_type_qf, 
         -average_length_of_episode_qf, -average_length_of_stay_qf, -location_qf,
         -admission_type_qf) %>% 
  mutate(hb = ifelse(hb == "S08000015", "Ayrshire and Arran", hb),
         hb = ifelse(hb == "S08000016", "Borders", hb),
         hb = ifelse(hb == "S08000017", "Dumfries and Galloway", hb),
         hb = ifelse(hb == "S08000018", "Fife", hb),
         hb = ifelse(hb == "S08000019", "Forth Valley", hb),
         hb = ifelse(hb == "S08000020", "Grampian", hb),
         hb = ifelse(hb == "S08000021", "Greater Glasgow and Clyde", hb),
         hb = ifelse(hb == "S08000022", "Highland", hb),
         hb = ifelse(hb == "S08000023", "Lanarkshire", hb),
         hb = ifelse(hb == "S08000024", "Lothian", hb),
         hb = ifelse(hb == "S08000025", "Orkney", hb),
         hb = ifelse(hb == "S08000026", "Shetland", hb),
         hb = ifelse(hb == "S08000027", "Tayside", hb),
         hb = ifelse(hb == "S08000028", "Western Isles", hb),
         hb = ifelse(hb == "S08000029", "Fife", hb),
         hb = ifelse(hb == "S08000030", "Tayside", hb),
         hb = ifelse(hb == "S08000031", "Greater Glasgow and Clyde", hb),
         hb = ifelse(hb == "S08000032", "Lanarkshire", hb),
         hb = ifelse(hb == "S92000003", "All of Scotland", hb),
         hb = ifelse(hb == "RA2701", "No Fixed Abode", hb),
         hb = ifelse(hb == "RA2702", "Rest of the UK", hb),
         hb = ifelse(hb == "RA2703", "Outside the UK", hb),
         hb = ifelse(hb == "RA2704", "Unknown Residency", hb),
         hb = ifelse(hb == "S27000001", "Non-NHS Provider", hb),
         hb = ifelse(hb == "SB0801", "The Golden Jubilee National Hospital", hb),
         hb = ifelse(hb == "SN0811", "National Facility NHS Louisa Jordan", hb),
           ) %>%
  rename(nhs_health_board = hb) %>% 
  separate(quarter,into = c("year", "quarter"), sep = "Q" ) %>% 
  mutate(pre_post_covid = ifelse(year %in% c("2017", "2018", "2019"), "pre", "post"))



```

# mean num episodes/quarter

```{r}
#mean_number of episodes

clean_hosp_admissions %>% 
  distinct(nhs_health_board)

clean_hosp_admissions %>% 
  filter(admission_type == "All Inpatients",
  nhs_health_board != "All of Scotland") %>% 
# nhs_health_board != "No Fixed Abode",
# nhs_health_board != "Rest of the UK",
# nhs_health_board != "Unknown Residency",
# nhs_health_board != "Outside the UK") %>% 
  arrange(episodes) %>% 
  group_by(year, quarter) %>% 
  summarise(av_episodes = mean(episodes),
            sd_episodes = sd(episodes),
            min_episodes = min(episodes),
            max_episdoes = max(episodes)) %>% 
  ggplot() +
  aes(x = quarter, y = av_episodes) +
  geom_point(size = 4, shape = 17, aes(colour = year)) +
  geom_line(aes(group = year, colour = year), size = 1) +
  geom_errorbar(aes(ymin = av_episodes - sd_episodes, 
                    ymax = av_episodes + sd_episodes,
                    width = 0.25, colour = year)) +
  stat_summary(fun.y = "mean", geom = "point", shape = 19, size = 4, color = "black",  
               aes(shape="")) +
  stat_summary(fun.y = mean, color = "black", geom = "line", aes(group = 1), 
               size = 1) + 
  # ylim(150000, 350000) +
  theme_bw() +
  scale_colour_brewer(palette = "Dark2") +
  geom_label(
    label = "Winter/Spring\n (Jan-Mar)", 
    x = 1,
    y = 325000,
    label.padding = unit(0.15, "lines"), 
    label.size = 0.15,
    color = "black"
  ) +
  geom_label(
    label = "Spring/Summer\n (Apr-Sep)", 
    x = 2.5,
    y = 325000,
    label.padding = unit(0.15, "lines"),
    label.size = 0.15,
    color = "black"
  ) +
  geom_label(
    label = "Autumn/Winter\n (Oct-Dec)", 
    x = 4,
    y = 325000,
    label.padding = unit(0.15, "lines"),
    label.size = 0.15,
    color = "black"
  ) +
  geom_vline(xintercept = 1.5, linetype = "dashed") +
  geom_vline(xintercept = 3.5, linetype = "dashed") +
   labs(
    title = "Mean Number of Inpatient Episodes per Quarter",
    subtitle = "Quarterly Data from Q3 2017-Q3 2022\n",
    x = "Quarter",
    y = "Inpatient Episodes")
 
```




#total num episdoes /quarter
```{r}

#total_number of episodes
clean_hosp_admissions %>% 
  filter(admission_type == "All Inpatients",
  nhs_health_board %in% "All of Scotland") %>%
  group_by(pre_post_covid, year, quarter) %>% 
  summarise(total_episodes = sum(episodes)) %>% 
  ggplot() +
  aes(x = quarter, y = total_episodes) +
  geom_point(size = 4, shape = 17, aes(colour = year)) +
  geom_line(aes(group = year, colour = year), size = 1) +
  stat_summary(fun.y = "mean", geom = "point", shape = 19, size = 4, color = "black",  aes(shape="")) +
  stat_summary(fun.y = mean, color = "black", geom = "line", aes(group = 1), size = 1) + 
  # ylim(500000, 18000) +
  theme_bw() +
  scale_colour_brewer(palette = "Dark2") +
  geom_label(
    label = "Winter/Spring\n (Jan-Mar)", 
    x = 1,
    y = 295000,
    label.padding = unit(0.15, "lines"), 
    label.size = 0.15,
    color = "black"
  ) +
  geom_label(
    label = "Spring/Summer\n (Apr-Sep)", 
    x = 2.5,
    y = 295000,
    label.padding = unit(0.15, "lines"),
    label.size = 0.15,
    color = "black"
  ) +
  geom_label(
    label = "Autumn/Winter\n (Oct-Dec)", 
    x = 4,
    y = 295000,
    label.padding = unit(0.15, "lines"),
    label.size = 0.15,
    color = "black"
  ) +
  geom_vline(xintercept = 1.5, linetype = "dashed") +
  geom_vline(xintercept = 3.5, linetype = "dashed") +
   labs(
    title = "Average Number of Inpatient Episodes per Month",
    subtitle = "Quarterly Data from Q3 2017-Q3 2022\n",
    x = "Quarter",
    y = "Inpatient Episodes") +
  ylim(200000, 300000)
 
```

# av stay per month all inpatients/quarter

```{r}
#av stay per month per quarter for all inpatients

clean_hosp_admissions %>% 
  filter(admission_type == "All Inpatients",
  nhs_health_board %in% "All of Scotland") %>%
  mutate(pre_post_covid = ifelse(year == c("2017", "2018", "2019"), "pre", "post")) %>% 
  group_by(pre_post_covid, year, quarter) %>% 
  summarise(av_stay = mean(stays)) %>% 
  ggplot() +
  aes(x = quarter, y = av_stay) +
  geom_point(size = 4, shape = 17, aes(colour = year)) +
  geom_line(aes(group = year, colour = year), size = 1) +
  stat_summary(fun.y = "mean", geom = "point", shape = 19, size = 4, color = "black") +
  scale_shape_manual("mean", values = "" ) +
  stat_summary(fun.y = mean, color = "black", geom = "line", aes(group = 1), size = 1) + 
  ylim(100000, 180000) +
  theme_bw() +
  scale_colour_brewer(palette = "Dark2") +
  geom_label(
    label = "Winter/Spring", 
    x = 1,
    y = 180000,
    label.padding = unit(0.1, "lines"), 
    label.size = 0.1,
    color = "black"
  ) +
  geom_label(
    label = "Spring / Summer", 
    x = 2.5,
    y = 180000,
    label.padding = unit(0.1, "lines"),
    label.size = 0.1,
    color = "black"
  ) +
  geom_label(
    label = "Autumn/Winter", 
    x = 4,
    y = 180000,
    label.padding = unit(0.1, "lines"),
    label.size = 0.1,
    color = "black"
  ) +
  geom_vline(xintercept = 1.5, linetype = "dashed") +
  geom_vline(xintercept = 3.5, linetype = "dashed") +
  labs(
    title = "Total Number of Inpatient Stays per Month",
    subtitle = "Quarterly Data from Q3 2017-Q3 2022\n",
    x = "Quarter",
    y = "Inpatient Stays")
```



```{r}
#av stays per month per quarter for all inpatients, pre/post covid
clean_hosp_admissions %>% 
  filter(admission_type == "All Inpatients",
  nhs_health_board %in% "All of Scotland") %>% 
  group_by(pre_post_covid, quarter) %>% 
  summarise(av_stay = mean(stays)) %>% 
  ggplot() +
  aes(x = quarter, y = av_stay) +
  geom_point(size = 4, shape = 17, aes(colour = pre_post_covid)) +
  geom_line(aes(group = pre_post_covid, colour = pre_post_covid), size = 1) +
  stat_summary(fun.y = "mean", geom = "point", shape = 19, size = 4, color = "black") +
  stat_summary(fun.y = mean, color = "black", geom = "line", aes(group = 1), size = 1) + 
  ylim(120000, 180000) +
  theme_bw() +
  scale_colour_brewer(palette = "Dark2") +
  geom_label(
    label = "Winter/Spring", 
    x = 1,
    y = 10000,
    label.padding = unit(0.1, "lines"), 
    label.size = 0.1,
    color = "black"
  ) +
  geom_label(
    label = "Spring / Summer", 
    x = 2.5,
    y = 10000,
    label.padding = unit(0.1, "lines"),
    label.size = 0.1,
    color = "black"
  ) +
  geom_label(
    label = "Autumn/Winter", 
    x = 4,
    y = 10000,
    label.padding = unit(0.1, "lines"),
    label.size = 0.1,
    color = "black"
  ) +
  geom_vline(xintercept = 1.5, linetype = "dashed") +
  geom_vline(xintercept = 3.5, linetype = "dashed") +
  labs(
    title = "Mean Number of Inpatient Stays per Month",
    subtitle = "Quarterly Data from Q3 2017-Q3 2022\n",
    x = "Quarter",
    y = "Inpatient Stays")
```
#av eps per month per quarter for all inpatients
```{r}
#av eps per month per quarter for all inpatients
clean_hosp_admissions %>% 
  filter(admission_type == "All Inpatients",
  nhs_health_board %in% "All of Scotland") %>% 
  group_by(pre_post_covid, quarter) %>% 
  summarise(av_episodes = mean(episodes),
            sd_episodes = sd(episodes)) %>% 
  ggplot() +
  aes(x = quarter, y = av_episodes) +
  geom_point(size = 4, shape = 17, aes(colour = pre_post_covid)) +
  geom_line(aes(group = pre_post_covid, colour = pre_post_covid), size = 1) +
  stat_summary(fun.y = "mean", geom = "point", shape = 19, size = 4, color = "black") +
  stat_summary(fun.y = mean, color = "black", geom = "line", aes(group = 1), size = 1) + 
  geom_errorbar(aes(ymin = av_episodes - sd_episodes, 
                     ymax = av_episodes + sd_episodes,
                     width = 0.25, colour = pre_post_covid)) +
  ylim(220000, 280000) +
  theme_bw() +
  scale_colour_brewer(palette = "Dark2") +
  geom_label(
    label = "Winter/Spring\n (Jan-Mar)", 
    x = 1,
    y = 17750,
    label.padding = unit(0.15, "lines"), 
    label.size = 0.15,
    color = "black"
  ) +
  geom_label(
    label = "Spring/Summer\n (Apr-Sep)", 
    x = 2.5,
    y = 17750,
    label.padding = unit(0.15, "lines"),
    label.size = 0.15,
    color = "black"
  ) +
  geom_label(
    label = "Autumn/Winter\n (Oct-Dec)", 
    x = 4,
    y = 17750,
    label.padding = unit(0.15, "lines"),
    label.size = 0.15,
    color = "black"
  ) +
  geom_vline(xintercept = 1.5, linetype = "dashed") +
  geom_vline(xintercept = 3.5, linetype = "dashed") +
   labs(
    title = "Average Number of Inpatient Episodes per Month",
    subtitle = "Quarterly Data from Q3 2017-Q3 2022\n",
    x = "Quarter",
    y = "Inpatient Episodes")
 
```


# episode number in emergency inpatients
```{r}
# episode number in emergency inpatients

clean_hosp_admissions %>% 
  filter(admission_type == "Emergency Inpatients") %>%
  group_by(pre_post_covid, quarter, year) %>% 
  summarise(av_episodes = mean(episodes)) %>% 
   ggplot() +
  aes(x = quarter, y = av_episodes) +
  geom_point(size = 4, shape = 17, aes(colour = year)) +
  geom_line(aes(group = year, colour = year), size = 1) +
  stat_summary(fun.y = "mean", geom = "point", shape = 19, size = 4, color = "black",  aes(shape="")) +
  stat_summary(fun.y = mean, color = "black", geom = "line", aes(group = 1), size = 1) + 
  ylim(5000, 9000) +
  theme_bw() +
  scale_colour_brewer(palette = "Dark2") +
  geom_label(
    label = "Winter/Spring\n (Jan-Mar)", 
    x = 1,
    y = 8750,
    label.padding = unit(0.15, "lines"), 
    label.size = 0.15,
    color = "black"
  ) +
  geom_label(
    label = "Spring/Summer\n (Apr-Sep)", 
    x = 2.5,
    y = 8750,
    label.padding = unit(0.15, "lines"),
    label.size = 0.15,
    color = "black"
  ) +
  geom_label(
    label = "Autumn/Winter\n (Oct-Dec)", 
    x = 4,
    y = 8750,
    label.padding = unit(0.15, "lines"),
    label.size = 0.15,
    color = "black"
  ) +
  geom_vline(xintercept = 1.5, linetype = "dashed") +
  geom_vline(xintercept = 3.5, linetype = "dashed") +
   labs(
    title = "Average Number of Emergency Inpatient Episodes per Month",
    subtitle = "Quarterly Data from Q3 2017-Q3 2022\n",
    x = "Quarter",
    y = "Inpatient Episodes")


# factor(quarter, ordered = TRUE, levels = c("3", "4", "1", "2")),
  

```

# average episode number in emergency inpatients pre and post vid
```{r}

# average episode number in emergency inpatients pre and post vid
clean_hosp_admissions %>% 
  filter(admission_type == "Emergency Inpatients") %>%
  group_by(pre_post_covid, quarter) %>% 
  summarise(av_episodes = mean(episodes)) %>% 
  ggplot() +
  aes(x = quarter, y = av_episodes) +
  geom_point(size = 4, shape = 17, aes(colour = pre_post_covid)) +
  geom_line(aes(group = pre_post_covid, colour = pre_post_covid), size = 1) +
  stat_summary(fun.y = "mean", geom = "point", shape = 19, size = 4, color = "black") +
  stat_summary(fun.y = mean, color = "black", geom = "line", aes(group = 1), size = 1) + 
  ylim(6000, 8000) +
  theme_bw() +
  scale_colour_brewer(palette = "Dark2") +
  geom_label(
    label = "Winter/Spring\n (Jan-Mar)", 
    x = 1,
    y = 7900,
    label.padding = unit(0.15, "lines"), 
    label.size = 0.15,
    color = "black"
  ) +
  geom_label(
    label = "Spring/Summer\n (Apr-Sep)", 
    x = 2.5,
    y = 7900,
    label.padding = unit(0.15, "lines"),
    label.size = 0.15,
    color = "black"
  ) +
  geom_label(
    label = "Autumn/Winter\n (Oct-Dec)", 
    x = 4,
    y = 7900,
    label.padding = unit(0.15, "lines"),
    label.size = 0.15,
    color = "black"
  ) +
  geom_vline(xintercept = 1.5, linetype = "dashed") +
  geom_vline(xintercept = 3.5, linetype = "dashed") +
   labs(
    title = "Average Number of Emergency Inpatient Episodes per Month",
    subtitle = "Quarterly Data from Q3 2017-Q3 2022\n",
    x = "Quarter",
    y = "Inpatient Episodes")
```
# Q3 first
# mean episode number in emergency inpatients pre and post vid
```{r}
# Q3 first
# mean episode number in emergency inpatients pre and post vid
clean_hosp_admissions %>% 
  filter(admission_type == "Emergency Inpatients",
         nhs_health_board == "All of Scotland") %>%
  group_by(pre_post_covid, quarter) %>% 
  summarise(av_episodes = mean(episodes),
            sd_episdoes = sd(episodes)) %>% 
  ggplot() +
  aes(x = factor(quarter, ordered = TRUE, levels = c("3", "4", "1", "2")), y = av_episodes) +
  geom_point(size = 4, shape = 17, aes(colour = pre_post_covid)) +
  geom_line(aes(group = pre_post_covid, colour = pre_post_covid), size = 1) +
  stat_summary(fun.y = "mean", geom = "point", shape = 19, size = 4, color = "black") +
  stat_summary(fun.y = mean, color = "black", geom = "line", aes(group = 1), size = 1) + 
  ylim(90000, 160000) +
  theme_bw() +
  scale_colour_brewer(palette = "Dark2") +
  geom_label(
    label = "Spring/Summer\n (Jul-Sep)", 
    x = 1,
    y = 150000,
    label.padding = unit(0.15, "lines"), 
    label.size = 0.15,
    color = "black"
  ) +
  geom_label(
    label = "Autumn/Winter\n (Oct-Mar)", 
    x = 2.5,
    y = 150000,
    label.padding = unit(0.15, "lines"),
    label.size = 0.15,
    color = "black"
  ) +
  geom_label(
    label = "Spring/Summer\n (Apr-June)", 
    x = 4,
    y = 150000,
    label.padding = unit(0.15, "lines"),
    label.size = 0.15,
    color = "black"
  ) +
  geom_vline(xintercept = 1.5, linetype = "dashed") +
  geom_vline(xintercept = 3.5, linetype = "dashed") +
   labs(
    title = "Average Number of Emergency Inpatient Episodes per Month",
    subtitle = "Quarterly Data from Q3 2017-Q3 2022\n",
    x = "Quarter",
    y = "Inpatient Episodes")





```
# Q3 first
# total episode number in emergency inpatients pre and post vid
```{r}
# Q3 first
# total episode number in emergency inpatients pre and post vid
clean_hosp_admissions %>% 
  filter(admission_type == "Emergency Inpatients") %>%
  group_by(pre_post_covid, quarter) %>% 
  summarise(total_episodes = sum(episodes)) %>% 
  ggplot() +
  aes(x = factor(quarter, ordered = TRUE, levels = c("3", "4", "1", "2")), y = total_episodes) +
  geom_point(size = 4, shape = 17, aes(colour = pre_post_covid)) +
  geom_line(aes(group = pre_post_covid, colour = pre_post_covid), size = 1) +
  # stat_summary(fun.y = "mean", geom = "point", shape = 19, size = 4, color = "black") +
  # stat_summary(fun.y = mean, color = "black", geom = "line", aes(group = 1), size = 1) + 
  theme_bw() +
  scale_colour_brewer(palette = "Dark2") +
  geom_label(
    label = "Spring/Summer\n (Jul-Sep)", 
    x = 1,
    y = 1250000,
    label.padding = unit(0.15, "lines"), 
    label.size = 0.15,
    color = "black"
  ) +
  geom_label(
    label = "Autumn/Winter\n (Oct-Mar)", 
    x = 2.5,
    y = 1250000,
    label.padding = unit(0.15, "lines"),
    label.size = 0.15,
    color = "black"
  ) +
  geom_label(
    label = "Spring/Summer\n (Apr-June)", 
    x = 4,
    y = 1250000,
    label.padding = unit(0.15, "lines"),
    label.size = 0.15,
    color = "black"
  ) +
  geom_vline(xintercept = 1.5, linetype = "dashed") +
  geom_vline(xintercept = 3.5, linetype = "dashed") +
   labs(
    title = "Total Number of Inpatient Episodes per Quarter",
    subtitle = "Quarterly Data from Q3 2017-Q3 2022\n",
    x = "Quarter",
    y = "Inpatient Episodes") +
  ylim(600000, 1300000)

```


```{r}
clean_hosp_admissions %>% 
  filter(admission_type == "All Inpatients") %>% 
  group_by(pre_post_covid, quarter) %>% 
  summarise(av_episodes = mean(episodes)) %>% 
  ggplot() +
  aes(x = factor(quarter, ordered = TRUE, levels = c("3", "4", "1", "2")), y = av_episodes) +
  geom_col()
  # geom_line(aes(group = pre_post_covid, colour = pre_post_covid), size = 1) +
  # stat_summary(fun.y = "mean", geom = "point", shape = 19, size = 4, color = "black") +
  # stat_summary(fun.y = mean, color = "black", geom = "line", aes(group = 1), size = 1) + 
  # theme_bw() +
  # scale_colour_brewer(palette = "Dark2") +
  # facet_wrap(~admission_type) +
  # geom_label(
  #   label = "Spring/Summer\n (Apr-Sep)", 
  #   x = 1,
  #   y = 7900,
  #   label.padding = unit(0.15, "lines"), 
  #   label.size = 0.15,
  #   color = "black"
  # ) +
  # geom_label(
  #   label = "Autumn/Winter\n (Jan-Mar)", 
  #   x = 2.5,
  #   y = 7900,
  #   label.padding = unit(0.15, "lines"),
  #   label.size = 0.15,
  #   color = "black"
  # ) +
  # geom_label(
  #   label = "Spring/Summer\n (Apr-Sep)", 
  #   x = 4,
  #   y = 7900,
  #   label.padding = unit(0.15, "lines"),
  #   label.size = 0.15,
  #   color = "black"
  # ) +
  # geom_vline(xintercept = 1.5, linetype = "dashed") +
  # geom_vline(xintercept = 3.5, linetype = "dashed") +
  #  labs(
  #   title = "Average Number of Emergency Inpatient Episodes per Month",
  #   subtitle = "Quarterly Data from Q3 2017-Q3 2022\n",
  #   x = "Quarter",
  #   y = "Inpatient Episodes")
```

#cleaning no separate quarters
```{r}
clean_hosp_admissions_qyear <- hospital_admissions %>% 
  clean_names() %>% 
  select(- quarter_qf, - hbqf, -location_qf, -admission_type_qf, 
         -average_length_of_episode_qf, -average_length_of_stay_qf, -location_qf,
         -admission_type_qf) %>% 
  mutate(hb = ifelse(hb == "S08000015", "Ayrshire and Arran", hb),
         hb = ifelse(hb == "S08000016", "Borders", hb),
         hb = ifelse(hb == "S08000017", "Dumfries and Galloway", hb),
         hb = ifelse(hb == "S08000018", "Fife", hb),
         hb = ifelse(hb == "S08000019", "Forth Valley", hb),
         hb = ifelse(hb == "S08000020", "Grampian", hb),
         hb = ifelse(hb == "S08000021", "Greater Glasgow and Clyde", hb),
         hb = ifelse(hb == "S08000022", "Highland", hb),
         hb = ifelse(hb == "S08000023", "Lanarkshire", hb),
         hb = ifelse(hb == "S08000024", "Lothian", hb),
         hb = ifelse(hb == "S08000025", "Orkney", hb),
         hb = ifelse(hb == "S08000026", "Shetland", hb),
         hb = ifelse(hb == "S08000027", "Tayside", hb),
         hb = ifelse(hb == "S08000028", "Western Isles", hb),
         hb = ifelse(hb == "S08000029", "Fife", hb),
         hb = ifelse(hb == "S08000030", "Tayside", hb),
         hb = ifelse(hb == "S08000031", "Greater Glasgow and Clyde", hb),
         hb = ifelse(hb == "S08000032", "Lanarkshire", hb),
         hb = ifelse(hb == "S92000003", "All of Scotland", hb),
         hb = ifelse(hb == "RA2701", "No Fixed Abode", hb),
         hb = ifelse(hb == "RA2702", "Rest of the UK", hb),
         hb = ifelse(hb == "RA2703", "Outside the UK", hb),
         hb = ifelse(hb == "RA2704", "Unknown Residency", hb),
         hb = ifelse(hb == "S27000001", "Non-NHS Provider", hb),
         hb = ifelse(hb == "SB0801", "The Golden Jubilee National Hospital", hb),
         hb = ifelse(hb == "SN0811", "National Facility NHS Louisa Jordan", hb),
           ) %>%
  rename(nhs_health_board = hb) 
  # separate(quarter,into = c("year", "quarter"), sep = "Q" ) %>% 
  # mutate(pre_post_covid = ifelse(year %in% c("2017", "2018", "2019"), "pre", "post"))
```


```{r}

#total number of inpatient episodes over time NOTE: do lockdowns affect this??
clean_hosp_admissions_qyear %>% 
filter(admission_type == "Emergency Inpatients",
  nhs_health_board %in% "All of Scotland") %>%
  group_by(quarter) %>% 
  summarise(total_episodes = sum(episodes)) %>% 
  ggplot() +
  aes(x = quarter, y = total_episodes) +
  geom_line(aes(group = 1, colour = "red")) +
  geom_point(size = 4, shape = 17, colour = "red") +
  geom_line(aes(group = quarter)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.5)) +
  scale_colour_brewer(palette = "Dark2") +
  geom_label(
    label = "Pre-Pandemic",
    x = 2.5,
    y = 140000,
    label.padding = unit(0.15, "lines"),
    label.size = 0.15,
    color = "black"
  ) +
  geom_label(
    label = "Pandemic",
    x = 20,
    y = 140000,
    label.padding = unit(0.15, "lines"),
    label.size = 0.15,
    color = "black"
  ) +
  geom_vline(xintercept = 11.5, linetype = "dashed") +
  # geom_vline(xintercept = 3.5, linetype = "dashed") +
   labs(
    title = "Total Number of Inpatient Episodes",
    subtitle = "Quarterly Data from Q3 2017-Q3 2022\n",
    x = "Quarter",
    y = "Inpatient Episodes") 
  # ylim(90000, 140000)
```


```{r}
#mean number of inpatient episodes over time NOTE: do lockdowns affect this??
#IGNORE THIS FOR NOW
clean_hosp_admissions%>% 
filter(admission_type == "Emergency Inpatients",
  nhs_health_board %in% "All of Scotland") %>%
  group_by(quarter) %>% 
  summarise(mean_episodes = mean(episodes),
            sd_episodes = sd(episodes)) %>% 
  ggplot() +
  aes(x = quarter, y = mean_episodes) +
  geom_line(aes(group = 1, colour = "red")) +
  geom_point(size = 4, shape = 17, colour = "red") +
  geom_line(aes(group = quarter)) +
  geom_errorbar(mapping = aes(ymin = mean_episodes - sd_episodes, 
                              ymax = mean_episodes + sd_episodes,
                              width = 0.25),
                colour = "red") +
  theme_bw() +
  # theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5)) +
  # scale_colour_brewer(palette = "Dark2") +
  geom_label(
    label = "Pre-2020",
    x = 2,
    y = 140000,
    label.padding = unit(0.15, "lines"),
    label.size = 0.15,
    color = "black"
  ) +
  geom_label(
    label = "Post-2020",
    x = 3,
    y = 140000,
    label.padding = unit(0.15, "lines"),
    label.size = 0.15,
    color = "black"
  ) +
  geom_vline(xintercept = 2.5, linetype = "dashed") +
  # geom_vline(xintercept = 3.5, linetype = "dashed") +
   labs(
    title = "Mean Number of Inpatient Episodes",
    subtitle = "Quarterly Data from Q3 2017-Q3 2022\n",
    x = "Quarter",
    y = "Inpatient Episodes") +
  ylim(100000, 140000)
```

```{r}
# anova test

summary(clean_hosp_admissions)

one_way <- aov(quarter ~ mean_episodes, sum_hosp_ad) # this does anova test

#create table to be tested
sum_hosp_ad <- clean_hosp_admissions%>% 
filter(admission_type == "Emergency Inpatients",
  nhs_health_board %in% "All of Scotland") %>%
  group_by(quarter) %>% 
  summarise(mean_episodes = mean(episodes),
            sd_episodes = sd(episodes))

summary(one_way)

## shows results of anova


```

#total number of emergency inpatient episodes over time
```{r}

#total number of emergency inpatient episodes over time
clean_hosp_admissions_qyear %>% 
filter(admission_type == "Emergency Inpatients",
  nhs_health_board %in% "All of Scotland") %>%
  group_by(quarter) %>% 
  summarise(total_episodes = sum(episodes)) %>% 
  ggplot() +
  aes(x = quarter, y = total_episodes) +
  geom_line(aes(group = 1, colour = "red")) +
  geom_point(size = 4, shape = 17, colour = "red") +
  geom_line(aes(group = quarter)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.5)) +
  scale_colour_brewer(palette = "Dark2") +
  geom_label(
    label = "Pre-Pandemic",
    x = 2.5,
    y = 140000,
    label.padding = unit(0.15, "lines"),
    label.size = 0.15,
    color = "black"
  ) +
  geom_label(
    label = "Pandemic",
    x = 20,
    y = 140000,
    label.padding = unit(0.15, "lines"),
    label.size = 0.15,
    color = "black"
  ) +
  geom_vline(xintercept = 11.5, linetype = "dashed") +
  # geom_vline(xintercept = 3.5, linetype = "dashed") +
   labs(
    title = "Total Number of Inpatient Episodes",
    subtitle = "Quarterly Data from Q3 2017-Q3 2022\n",
    x = "Quarter",
    y = "Inpatient Episodes") +
  ylim(80000, 140000)

```

#speciality cleaning script
```{r}
clean_hospital_admissions_speciality <- hospital_admissions_speciality %>% 
  clean_names() %>% 
  select(- quarter_qf, - hbqf, -location_qf, -admission_type_qf, 
         -average_length_of_episode_qf) %>% 
  mutate(hb = ifelse(hb == "S08000015", "Ayrshire and Arran", hb),
         hb = ifelse(hb == "S08000016", "Borders", hb),
         hb = ifelse(hb == "S08000017", "Dumfries and Galloway", hb),
         hb = ifelse(hb == "S08000018", "Fife", hb),
         hb = ifelse(hb == "S08000019", "Forth Valley", hb),
         hb = ifelse(hb == "S08000020", "Grampian", hb),
         hb = ifelse(hb == "S08000021", "Greater Glasgow and Clyde", hb),
         hb = ifelse(hb == "S08000022", "Highland", hb),
         hb = ifelse(hb == "S08000023", "Lanarkshire", hb),
         hb = ifelse(hb == "S08000024", "Lothian", hb),
         hb = ifelse(hb == "S08000025", "Orkney", hb),
         hb = ifelse(hb == "S08000026", "Shetland", hb),
         hb = ifelse(hb == "S08000027", "Tayside", hb),
         hb = ifelse(hb == "S08000028", "Western Isles", hb),
         hb = ifelse(hb == "S08000029", "Fife", hb),
         hb = ifelse(hb == "S08000030", "Tayside", hb),
         hb = ifelse(hb == "S08000031", "Greater Glasgow and Clyde", hb),
         hb = ifelse(hb == "S08000032", "Lanarkshire", hb),
         hb = ifelse(hb == "S92000003", "All of Scotland", hb),
         hb = ifelse(hb == "RA2701", "No Fixed Abode", hb),
         hb = ifelse(hb == "RA2702", "Rest of the UK", hb),
         hb = ifelse(hb == "RA2703", "Outside the UK", hb),
         hb = ifelse(hb == "RA2704", "Unknown Residency", hb)
           ) %>%
  rename(nhs_health_board = hb) %>% 
  separate(quarter,into = c("year", "quarter"), sep = "Q" ) %>% 
  mutate(pre_post_covid = ifelse(year %in% c("2017", "2018", "2019"), "pre", "post")) 


clean_hospital_admissions_speciality %>% 
  distinct(specialty_name) 
  


```


# Mean Inpatient Episodes Pre- and During Pandemic
```{r}
clean_hosp_admissions %>% 
  filter(admission_type == "Emergency Inpatients",
  nhs_health_board %in% "All of Scotland") %>%
  group_by(pre_post_covid, quarter) %>% 
  summarise(av_episodes = mean(episodes),
            sd_av_episodes = sd(episodes),
            min_av_episodes = min(episodes),
            max_episodes = max(episodes)) %>% 
  ggplot() +
  aes(x = factor(quarter, ordered = TRUE, levels = c("3", "4", "1", "2")), y = av_episodes) +
  geom_point(size = 4, shape = 17, aes(colour = pre_post_covid)) +
  geom_line(aes(group = pre_post_covid, colour = pre_post_covid), linewidth = 1) +
  # stat_summary(fun.y = "mean", geom = "point", shape = 19, size = 4, color = "black") +
  # stat_summary(fun.y = mean, color = "black", geom = "line", aes(group = 1), size = 1) +
  geom_errorbar(aes(ymin = av_episodes - sd_av_episodes, 
                              ymax = av_episodes + sd_av_episodes, 
                              colour = pre_post_covid,
                    width = 0.25)
                ) +
  theme_bw() +
  scale_colour_brewer(palette = "Dark2") +
  geom_label(
    label = "Pre-Pandemic",
    x = 0.87,
    y = 145000,
    label.padding = unit(0.15, "lines"),
    label.size = 0.15,
    color = "black"
  ) +
  geom_label(
    label = "Pandemic",
    x = 4.25,
    y = 145000,
    label.padding = unit(0.15, "lines"),
    label.size = 0.15,
    color = "black"
  ) +
  geom_vline(xintercept = 2.5, linetype = "dashed") +
  # geom_vline(xintercept = 3.5, linetype = "dashed") +
   labs(
    title = "Mean Inpatient Episodes Pre- and During Pandemic",
    subtitle = "Quarterly Data from Q3 2017-Q3 2022\n",
    x = "Quarter",
    y = "Inpatient Episodes") +
    ylim(100000, 145000)
  
```
```{r}

# H0: mean_inpatient_episodes are the same
# Ha: mean_inpatient_episodes are different

# anova test

summary(clean_hosp_admissions)

one_way <- aov(av_episodes ~ quarter, sum_hosp_ad_v2) # this does anova test

plot(episodes ~ quarter, sum_hosp_ad_v2)

#create table to be tested
sum_hosp_ad_v2 <- clean_hosp_admissions %>% 
  filter(admission_type == "Emergency Inpatients",
  nhs_health_board %in% "All of Scotland") %>%
  group_by(pre_post_covid, quarter) %>% 
  summarise(av_episodes = mean(episodes),
            sd_av_episodes = sd(episodes),
            min_av_episodes = min(episodes),
            max_episodes = max(episodes))

# shows results of anova
summary(one_way)


clean_hosp_admissions %>% 
  filter(admission_type == "Emergency Inpatients",
  nhs_health_board %in% "All of Scotland") %>%
  group_by(pre_post_covid, quarter) %>% 
  ggplot() +
  geom_col(aes(x = quarter, y = episodes))

  


```

